---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```
# DFplyr

<!-- badges: start -->
<!-- badges: end -->

The goal of DFplyr is to enable `dplyr` and `ggplot2` support for
`S4Vectors::DataFrame` by providing the appropriate extension methods. As row
names are an important feature of many Bioconductor structures, these are
preserved where possible.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("jonocarroll/DFplyr")
```

You can install from [Bioconductor](https://bioconductor.org) with:

``` r
if (!require("BiocManager", quietly =TRUE))
    install.packages("BiocManager")

# The following initializes usage of Bioc devel
BiocManager::install(version='devel')

BiocManager::install("DFplyr")
```

## Examples

First create an S4Vectors `DataFrame`, including S4 columns if desired

```{r}
suppressMessages(library(S4Vectors))
m <- mtcars[, c("cyl", "hp", "am", "gear", "disp")]
d <- as(m, "DataFrame")
d$grX <- GenomicRanges::GRanges("chrX", IRanges::IRanges(1:32, width = 10))
d$grY <- GenomicRanges::GRanges("chrY", IRanges::IRanges(1:32, width = 10))
d$nl <- IRanges::NumericList(lapply(d$gear, function(n) round(rnorm(n), 2)))
d
```

`DataFrame`s can then be used in `dplyr` calls the same as `data.frame` or
`tibble` objects. Support for working with S4 columns is enabled provided they
have appropriate functions. Adding multiple columns will result in the new
columns being created in alphabetical order

```{r}
suppressMessages(library(DFplyr))

mutate(d, newvar = cyl + hp)

mutate(d, nl2 = nl * 2)

mutate(d, length_nl = lengths(nl))

mutate(d,
    chr = GenomeInfoDb::seqnames(grX),
    strand_X = BiocGenerics::strand(grX),
    end_X = BiocGenerics::end(grX)
)
```

the object returned remains a standard `DataFrame`, and further calls can be 
piped with `%>%`


```{r}
mutate(d, newvar = cyl + hp) %>%
    pull(newvar)
```

Some of the variants of the `dplyr` verbs also work

```{r}
mutate_if(d, is.numeric, ~ .^2)

mutate_if(d, ~ inherits(., "GRanges"), BiocGenerics::start)
```

Use of `tidyselect` helpers is limited to within `dplyr::vars()` calls and using 
the `_at` variants

```{r}
mutate_at(d, vars(starts_with("c")), ~ .^2)

select_at(d, vars(starts_with("gr")))
```

Importantly, grouped operations are supported. `DataFrame` does not 
natively support groups (the same way that `data.frame` does not) so these
are implemented specifically for `DFplyr`

```{r}
group_by(d, cyl, am)
```

Other verbs are similarly implemented, and preserve row names where possible

```{r}
select(d, am, cyl)

arrange(d, desc(hp))

filter(d, am == 0)

slice(d, 3:6)

group_by(d, gear) %>%
    slice(1:2)
```

`rename` works in the {dplyr} sense of taking `new = old` replacements with NSE syntax

```{r}
select(d, am, cyl) %>%
    rename(foo = am)
```

Row names are not preserved when there may be duplicates or they don't make
sense, otherwise the first label (according to the current de-duplication
method, in the case of `distinct`, this is via `BiocGenerics::duplicated`). This
may have complications for S4 columns.

```{r}
distinct(d)

group_by(d, cyl, am) %>%
    tally(gear)

count(d, gear, am, cyl)
```

## Joins

Joins attempt to preserve rownames and grouping wherever possible

```{r}
Da <- as(starwars[, c("name", "eye_color", "height", "mass")], "DataFrame") |>
    head(10) |>
    group_by(eye_color)
Da

Db <- as(starwars[, c("name", "eye_color", "homeworld")], "DataFrame")
Db

left_join(Da, Db)

right_join(Da, Db)

inner_join(Da, Db[1:3, ])

full_join(Da, Db[1:3, ])
```

## Coverage

Most `dplyr` functions are implemented.

If you find any which are not, please [file an issue](https://github.com/jonocarroll/DFplyr/issues/new).
