---
title: "Example Usage"
author: 
  - name: Jonathan Carroll
    email: rpkg@jcarroll.com.au
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('DFplyr')`"
vignette: >
  %\VignetteIndexEntry{Example Usage}
  %\VignetteEncoding{UTF-8}  
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL
)
```

# Basics

## Install `DFplyr`

`r Biocpkg("DFplyr")` is a `R` package available via the 
[Bioconductor](http://bioconductor.org) repository for packages and can be
downloaded via `BiocManager::install()`:

```{r "install", eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("DFplyr")

## Check that you have a valid Bioconductor installation
BiocManager::valid()
```

## Background

`r Biocpkg("DFplyr")` is inspired by `r CRANpkg("dplyr")` which implements a
wide variety of common data manipulations (`mutate`, `select`, `filter`) but
which only operates on objects of class `data.frame` or `tibble` (from `r
CRANpkg("tibble")`).

When working with `r Biocpkg("S4Vectors")` `DataFrame`s - which are frequently
used as components of, for example `r Biocpkg("SummarizedExperiment")` objects -
a common workaround is to convert the `DataFrame` to a `tibble` in order to then
use `r CRANpkg("dplyr")` functions to manipulate the contents, before converting
back to a `DataFrame`.

This has several drawbacks, including the fact that `tibble` does not support 
rownames (and `r CRANpkg("dplyr")` frequently does not preserve them), does not 
support S4 columns (e.g. `r Biocpkg("IRanges")` vectors), and requires the back 
and forth transformation any time manipulation is desired.

# Quick start to using `DFplyr`

```{r "start", message=FALSE}
library("DFplyr")
```

To being with, we create an `r Biocpkg("S4Vectors")` `DataFrame`, including some
S4 columns

```{r "create_d", message=FALSE}
library(S4Vectors)
m <- mtcars[, c("cyl", "hp", "am", "gear", "disp")]
d <- as(m, "DataFrame")
d$grX <- GenomicRanges::GRanges("chrX", IRanges::IRanges(1:32, width = 10))
d$grY <- GenomicRanges::GRanges("chrY", IRanges::IRanges(1:32, width = 10))
d$nl <- IRanges::NumericList(lapply(d$gear, function(n) round(rnorm(n), 2)))
d
```

This will appear in RStudio's environment pane as a 

``` 
Formal class DataFrame (dplyr-compatible)
```

when using `r Biocpkg("DFplyr")`. No interference with the actual object is
required, but this helps identify that `r CRANpkg("dplyr")`-compatibility is
available.

`DataFrame`s can then be used in `r CRANpkg("dplyr")`-like calls the same as
`data.frame` or `tibble` objects. Support for working with S4 columns is enabled
provided they have appropriate functions. Adding multiple columns will result in
the new columns being created in alphabetical order. For example, adding a new 
column `newvar` which is the sum of the `cyl` and `hp` columns

```{r "mutate_newvar"}
mutate(d, newvar = cyl + hp)
```

or doubling the `nl` column as `nl2`

```{r "nl2"}
mutate(d, nl2 = nl * 2)
```

or calculating the `length()` of the `nl` column cells as `length_nl`

```{r "length_nl"}
mutate(d, length_nl = lengths(nl))
```

Transformations can involve S4-related functions, such as extracting the 
`seqnames()`, `strand()`, and `end()` of the `grX` column

```{r "s4cols"}
mutate(d,
    chr = GenomeInfoDb::seqnames(grX),
    strand_X = BiocGenerics::strand(grX),
    end_X = BiocGenerics::end(grX)
)
```

the object returned remains a standard `DataFrame`, and further calls can be 
piped with `%>%`, in this case extracting the newly created `newvar` column

```{r "pipe"}
mutate(d, newvar = cyl + hp) %>%
    pull(newvar)
```

Some of the variants of the `dplyr` verbs also work, such as transforming the
numeric columns using a quosure style lambda function, in this case squaring
them

```{r "mutate_if"}
mutate_if(d, is.numeric, ~ .^2)
```

or extracting the `start` of all of the `"GRanges"` columns

```{r "mutate_if_granges"}
mutate_if(d, ~ isa(., "GRanges"), BiocGenerics::start)
```

Use of `r CRANpkg("tidyselect")` helpers is limited to within `vars()`
calls and using the `_at` variants

```{r "at_mutate"}
mutate_at(d, vars(starts_with("c")), ~ .^2)
```

and also works with other verbs

```{r "at_select"}
select_at(d, vars(starts_with("gr")))
```

Importantly, grouped operations are supported. `DataFrame` does not 
natively support groups (the same way that `data.frame` does not) so these
are implemented specifically for `DFplyr` with group information shown at the 
top of the printed output

```{r "group_by"}
group_by(d, cyl, am)
```

Other verbs are similarly implemented, and preserve row names where possible.
For example, selecting a limited set of columns using non-standard evaluation
(NSE)

```{r "rownames"}
select(d, am, cyl)
```

Arranging rows according to the ordering of a column

```{r "rownames_arrange"}
arrange(d, desc(hp))
```

Filtering to only specific values appearing in a column

```{r "rownames_filter"}
filter(d, am == 0)
```

Selecting specific rows by index

```{r "rownames_slice"}
slice(d, 3:6)
```

These also work for grouped objects, and also preserve the rownames, e.g.
selecting the first two rows from _each group_ of `gear`

```{r "grouped_slice"}
group_by(d, gear) %>%
    slice(1:2)
```

`rename` is itself renamed to `rename2` due to conflicts between 
`r CRANpkg("dplyr")` and `r Biocpkg("S4Vectors")`, but works in the 
`r CRANpkg("dplyr")` sense of taking `new = old` replacements with NSE syntax

```{r "rename2"}
select(d, am, cyl) %>%
    rename2(foo = am)
```

Row names are not preserved when there may be duplicates or they don't make
sense, otherwise the first label (according to the current de-duplication
method, in the case of `distinct`, this is via `BiocGenerics::duplicated`). This
may have complications for S4 columns.

```{r "distinct"}
distinct(d)
```

Behaviours are ideally the same as those of `r CRANpkg("dplyr")` wherever
possible, for example a grouped tally

```{r "group_tally"}
group_by(d, cyl, am) %>%
    tally(gear)
```

or a count with weights

```{r "count"}
count(d, gear, am, cyl)
```

Joins attempt to preserve rownames and grouping wherever possible

```{r}
Da <- as(starwars[, c("name", "eye_color", "height", "mass")], "DataFrame") |> 
  head(10) |> 
  group_by(eye_color)
Da

Db <- as(starwars[, c("name", "eye_color", "homeworld")], "DataFrame")
Db

left_join(Da, Db)

right_join(Da, Db)

inner_join(Da, Db[1:3, ])

full_join(Da, Db[1:3, ])
```

## Citing `DFplyr`

We hope that `r Biocpkg("DFplyr")` will be useful for your research. Please use
the following information to cite the package and the overall approach. Thank
you!

```{r "citation"}
citation("DFplyr")
```


## Session Information.

```{r reproduce3, echo=FALSE}
options(width = 120)
sessioninfo::session_info()
```
